\texttt{\emph{Incertae sedis}}\texttt{ taxa in supertree construction}

### 2018-03-05 doc/incertae_sedis_latex2.tex
Update document for revised definition of i.s. exclude sets.

\subsubsection{Reduced exclude sets for \emph{incertae sedis} taxa}

@@

An \emph{incertae sedis} taxon can be moved into any of the descendants
of its siblings. Therefore, the exclude $\mathcal{E}(n)$ set for
a node $n$ contains the children of all the ancestors of $n$, unless
those children are incertae sedis. We can compute exclude sets recursively
by writing the exclude set of a node $n$ in terms of the exclude
set of its parent. If we use the terminology that the include and
exclude sets for a node $n$ are $\mathcal{I}(n)$ and $\mathcal{E}(n)$,
then we have 
\begin{align}
\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m)\big|m\in siblings(n),m\text{ not \emph{incertae} \emph{sedis}}\right],\label{eq:exclude-set-formula}
\end{align}
This leads to a pre-order recursion that terminates if the exclude
set for the root note is set to $\{\}$ as a boundary condition.

\subsubsection{Alternative formulation}

Consider an I.S. clade $A$ that is more rootward than an I.S. clade
$B$. With the above formulation, we could not place $A$ as sister
to $B$ and then place $B$ within $A$, because $A$ excludes $B$.
To solve this problem, we could modify formula (\ref{eq:exclude-set-formula})
above to avoid excluding \texttt{incertae\_sedis\_inherited} tips
of a sister taxon $m$. Thus, we could modify formula (\ref{eq:exclude-set-formula})
to refer to $\mathcal{I}^{\prime}(m)$ instead of $\mathcal{I}(m)$,
where 
\begin{align*}
\mathcal{I}^{\prime}(m) & =\begin{cases}
m & \text{if }m\text{ is a leaf}\\
\left[\mathcal{I}^{\prime}(c)\big|c\in children(m),c\text{ not \emph{incertae sedis}}\right] & \text{otherwise.}
\end{cases}
\end{align*}
This would not rule out placing $B$ as sister to $A$ and then placing
$A$ within $B$.

@@

Equation (\ref{eq:exclude-set-formula}) leads to the following algorithm
to compute the exclude set for all nodes in a tree.
\begin{enumerate}
\item Set the exclude set of the root node to be empty
\item For each \emph{node} (except the root) in preorder
\begin{itemize}
\item combine the \emph{exclude} set of the parent node with the \emph{include}
set of non-\emph{incertae-sedis} siblings.
\item store this set in a hash, with key \emph{node}
\end{itemize}
\end{enumerate}
This algorithm is currently implemented in \emph{otc-solve-subproblem}.
We store the sets as \emph{std::set}.

@@

-~Erysipelotrichaceae~broken
\end{lyxcode}
\end{comment}
\end{lyxcode}
Should we do this? We could say: 
\begin{itemize}
\item there are 331 incertae sedis taxa that are mentioned in input trees.
\begin{itemize}
\item what is the kingom / phylum / class / order / genus / species
\end{itemize}
\item we placed 14 incertae sedis taxa inside a sister taxon.
\item we confirmed $n$ incertae sedis taxa as being separate from all sister
(sampled) taxa??
\item we avoiding breaking 8 taxa that had IS taxa placed inside them.
\item we allowed $z_{1}$ new taxa into the synthesis tree that were incertae
sedis.
\item we allowed $z_{2}$ new taxa into the synthesis tree that are marked
as extinct.
\item some nodes have as many as $w$\emph{ incertae sedis} children, making
them unbrowseable when incertae sedis children are not excluded.
\item $v_{1}$ input trees were previously excluded \emph{entirely} because
they are nested within in an incertae sedis taxon.
\item $v_{2}$ input trees were previously excluded \emph{partially} because
they are nested within in an incertae sedis taxon.
\end{itemize}
Currently the numbers $z_{1}$ and $z_{2}$.

>>>>

\subsubsection{Partially reduced exclude sets for \emph{incertae sedis} taxa}

@@

An \emph{incertae sedis} taxon can be moved into any of the descendants
of its siblings. Therefore, the exclude $\mathcal{E}(n)$ set for
a node $n$ contains the children of all the ancestors of $n$, unless
those children are incertae sedis. We can compute exclude sets recursively
by writing the exclude set of a node $n$ in terms of the exclude
set of its parent. If we use the terminology that the include and
exclude sets for a node $n$ are $\mathcal{I}(n)$ and $\mathcal{E}(n)$,
then we have 
\begin{align}
\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m)\big|m\in siblings(n),m\text{ not \emph{incertae} \emph{sedis}}\right],\label{eq:exclude-set-formula-1}
\end{align}
This leads to a pre-order recursion that terminates if the exclude
set for the root note is set to $\{\}$ as a boundary condition.

\subsubsection{Extended formulation}

If an \emph{incertae sedis} taxon $A$ is placed within a sibling
clade, then it may gain a new \emph{incertae sedis} sibling $B$.
We could then possibly place $B$ within $A$. However, the above
definition does not allow this possibility. We therefore adjust the
above definition to exclude only descendants of sibling taxon only
the path to the descendant does not pass through a node labeled as
\emph{incertae sedis}. Such descendants of a sibling $m$ are given
by:
\begin{align*}
\mathcal{I}^{\prime}(m) & =\begin{cases}
m & \text{if }m\text{ is a leaf}\\
\left[\mathcal{I}^{\prime}(c)\big|c\in children(m),c\text{ not \emph{incertae sedis}}\right] & \text{otherwise.}
\end{cases}
\end{align*}
We then modify the formula above to exclude $\mathcal{I}^{\prime}(m)$
instead of $\mathcal{I}(m)$ for each sibling $m$:
\begin{align}
\mathcal{E}^{\prime}(n) & =\mathcal{E}^{\prime}(parent(n))\cup\left[\mathcal{I}^{\prime}(m)\big|m\in siblings(n),m\text{ not \emph{incertae} \emph{sedis}}\right],\label{eq:exclude-set-formula-2}
\end{align}

This less stringent formula is the one we will apply throughout this
paper.

@@

Equation (\ref{eq:exclude-set-formula-1}) leads to the following
algorithm to compute the exclude set for all nodes in a tree.
\begin{enumerate}
\item Set the exclude set of the root node to be empty
\item For each \emph{node} (except the root) in preorder
\begin{itemize}
\item combine the \emph{exclude} set of the parent node with the \emph{include}
set of non-\emph{incertae-sedis} siblings.
\item store this set in a hash, with key \emph{node}
\end{itemize}
\end{enumerate}
This algorithm is currently implemented in \emph{otc-solve-subproblem}.
We store the sets as \emph{std::set}.

@@

-~Erysipelotrichaceae~broken
\end{lyxcode}
\end{comment}
\end{lyxcode}
Should we do this? We could say: 
\begin{itemize}
\item there are 331 incertae sedis taxa that are mentioned in input trees.
\begin{itemize}
\item what is the kingom / phylum / class / order / genus / species
\end{itemize}
\item we placed 14 incertae sedis taxa inside a sister taxon. (``incertae
sedis'', ``versus ``unplaced''?)
\item we placed $o$ incertae sedis taxa \emph{outside} all sister taxa.
\item we confirmed $n$ incertae sedis taxa as being separate from all sister
(sampled) taxa??
\item we avoiding breaking 8 taxa that had IS taxa placed inside them.
\item we allowed $z_{1}$ new taxa into the synthesis tree that were incertae
sedis.
\item we allowed $z_{2}$ new taxa into the synthesis tree that are marked
as extinct.
\item some nodes have as many as $w$\emph{ incertae sedis} children, making
them unbrowseable when incertae sedis children are not excluded.
\item $v_{1}$ input trees were previously excluded \emph{entirely} because
they are nested within in an incertae sedis taxon.
\item $v_{2}$ input trees were previously excluded \emph{partially} because
they are nested within in an incertae sedis taxon.
\end{itemize}
Currently the numbers $z_{1}$ and $z_{2}$.
